---
title: "Discrete Probability Simulations in R"
author: "Francois de Ryckel"
date: "2023-11-14"
categories: [Probability, Simulations]
editor: source
date-modified: "2022-11-14"
---

The idea behind this post is to collect various numerical methods to simulate discrete probability problems. 

# Expectation of a uniform variable. 

Question: what is the expectation if one square a number that is picked at random out of a hat (with replacement) that contains the numbers 1 to 100.  

```{r}
expec <- mean(sample(1:100, size = 1000000, replace = TRUE)^2)
print(expec)
```

The calculated expectation should be: $$\sum_{x=1}^{100} x^2 P(X=x) = \sum_{x=1}^{100} x^2 \frac{1}{n} = \frac{101 \cdot 201}{6} = 3383.5$$

We could connect this to the [Jensen's inequality](../jensen-inequality/index.qmd) (as we are dealing with a convex function) and show that indeed the expectation of the function is greater than the function of the expectation: $\mathbb{E}[f(X)] \geq f(\mathbb{E}[X])$

```{r}
exp_square <- mean(sample(1:100, 1000000, replace = TRUE)^2)
square_exp <- (mean(sample(1:100, 1000000, replace = TRUE)))^2

print(exp_square)
print(square_exp)
```

