---
title: "05 - Trinomials models for Quantitative Finance"
author: "Francois de Ryckel"
date: "2023-04-08"
description: 'Creating a trinomial model and deriving the Forward Kolmogorov Equation.'
categories: [Probability, Probability Density Function]
editor: source
date-modified: "2023-07-23"
---


# Transition Density Function  


```{r}
#| warning: false
#| message: false

library(tibble)     # tibble()
library(dplyr)      # mutate()
library(ggplot2)    # ggplot()
```

## Setting the stage 

A discrete random variable (DRV) $y$ can either go up with a probability of $\alpha$ or down with a probability of $\alpha$ or stay the same with a probability of $1 - 2\alpha$.  

```{mermaid}
flowchart RL
  y' -- alpha --> y'+delta_y
  y' -- 1 - 2*alpha --> y
  y' -- alpha --> y'-delta_y
```

We call that a trinomial walk. 

## Generating an instance of a trinomial walk 

```{r}
alpha <- 0.3     # probability to go up or down 
                 # Hence, prob to stay the same is 0.4

#let's do a 252 steps trinomial walk (aka a year of daily movement). 

num_steps <- 252
prob <- runif(num_steps)

df <- tibble(step = 1:num_steps, prob = prob) |> 
  mutate(direction = if_else(prob < alpha, -1, if_else(prob > (1 - alpha), 1, 0)), 
         cum_walk = cumsum(direction))

ggplot(df, aes(x = step, y = cum_walk)) + 
  geom_line() + 
  ggtitle(label = 'Instance of a trinomial walk', subtitle = 'with 252 steps and symetric move') + 
  ylab(label = 'Cumulative distance')
```

This is just one instance of a trinomial walk.  In reality, we are interested in getting to know the probabilistic properties of the $y$ variable. 

# Deriving the Forward equation 

$$Prob(a<y'<b \text{ at time t' } | \text{ y at time t}) = \int_a^b p(y, t; y', t') dy'$$ {#eq-tran-prob-fct}

This (@eq-tran-prob-fct) means: *What is the probability that the random variable y' lies between a and b at time t' given it was at y at time t?* In this case (y, t) are given, they are constant, they are known; while (y', t') are the variables. 

We re-write this (@eq-tran-prob-fct) for conciseness as $P(y, t; y', t')$.  

Hence, another way to write (@eq-tran-prob-fct) is 
$$P(y, t; y', t') = \alpha \cdot P(y, t; y'+\delta y, t'-\delta t) + (1-2\alpha) P(y, t; y', t'-\delta t) + \alpha \cdot P(y, t; y'-\delta y, t' - \delta t)$$ {#eq-tran-prob-fct2}

Each terms in the sum of (@eq-tran-prob-fct2) could be evaluated using a *Taylor Series Expansion*.  Note that $\delta t^2 << \delta t$ as $\delta t$ is already quite small.  

$$P(y, t; y' + \delta y, t'-\delta t)  \approx P(y,t;y',t') + \delta y \frac{\partial P}{\partial y'} - \delta t \frac{\partial P}{\partial t'} + \frac{1}{2} \delta y^2 \frac{\partial^2 P}{\partial y'^2} + \dots$$
$$P(y, t; y', t'-\delta t)  \approx P(y,t;y',t') - \delta t \frac{\partial P}{\partial t'} + \dots$$

$$P(y, t; y'- \delta y, t'-\delta t)  \approx P(y,t;y',t') - \delta y \frac{\partial P}{\partial y'} - \delta t \frac{\partial P}{\partial t'} + \frac{1}{2} \delta y^2 \frac{\partial^2 P}{\partial y'^2} + \dots$$

We have ignored all the terms less than $\delta t$.

Adding the 3 equations above with their coefficients, we end up with 

$$\delta t \frac{\partial P}{\partial t'} = \alpha \delta y^2 \frac{\partial^2 P}{\partial y'^2}$$
$$\frac{\partial P}{\partial t'} = \alpha \frac{\delta y^2}{\delta t} \frac{\partial^2 P}{\partial y'^2}$$

Note how $\alpha$, $\delta t$ and $\delta y$ are all positive values.  Hence, we can let $C^2 = \alpha \frac{\delta y^2}{\delta t}$, and we get: 
$$\frac{\partial P}{\partial t'} = C^2 \frac{\partial^2 P}{\partial y'^2}$$ {#eq-FKE}

This last (@eq-FKE) can be recognized as the **Forward Kolmogorov Equation** or Heat-diffusion equation or also Fokker-Plank equation.  

Note that: 

* We used $P$ instead of $P(y, t; y', t')$ just for brievety
* This is a PDE for p with 2 independent variables $y'$ and $t'$
* $y$ and $t$ are like parameters.  They are fixed, they are starting point 
* This should model a random-walk that is finite in a finite time.

# Solving the FKE (by similarity reduction)

To solve this PDE, we solve it by (as per the CQF) similarity reduction.  We use a solution of the form 
$$P = t'^a f \left( \frac{y'}{t'^b} \right) \space a, b \in \mathbb{R}$$ {#eq-FKE-sol}

Letting $\xi = \frac{y'}{t'^b}$, we are looking for a solution of the form $$P = t'^a f(\xi) $$

Finding the partial derivatives based on the above solution's form.  

$$\frac{\partial P}{\partial y'} = t'^a \cdot \frac{df}{d \xi} \cdot \frac{\partial \xi}{\partial y'}$$ 
Note how f is just a function of $\xi$ while $\xi$ is a function of both $y'$ and $t'$; hence the difference in notation for the derivatives. 

Since $\frac{\partial \xi}{\partial y'} = t'^{-b}$, we have  $$\frac{\partial P}{\partial y'} = t'^{a-b} \cdot \frac{df}{d \xi} $$
$$\frac{\partial P^2}{\partial y'^2} = t'^{a-b} \frac{d^2f}{d \xi^2} \frac{\partial \xi}{\partial y'} = t'^{a-2b} \frac{d^2f}{d \xi^2}$$

Also, $\frac{\partial \xi}{\partial t'} = -b \cdot y' \cdot t'^{-b-1}$.  Using product rule to find $\frac{\partial P}{\partial t'}$, we get: 
$$\frac{\partial P}{\partial t'} = a t'^{a-1} f(\xi) + t'^a \frac{df}{d \xi} \frac{\partial \xi}{\partial t'} = a \cdot t'^{a-1} \cdot f(\xi) - b \cdot t'^{a-b-1} \cdot y' \cdot \frac{df}{d \xi} $$ which we could also re-write as: 
$$\frac{\partial P}{\partial t'} = a \cdot t'^{a-1} \cdot f(\xi) - b \cdot t'^{a-1} \cdot \xi \cdot \frac{df}{d \xi}$$ since $\xi = \frac{y'}{t'^b}$

Putting everything back together into @eq-FKE, we get: 
$$a \cdot t'^{a-1} \cdot f(\xi) - b \cdot t'^{a-1} \cdot \xi \cdot \frac{df}{d \xi} = C^2 \cdot t'^{a-2b} \frac{d^2f}{d \xi^2}$$ {#eq-FKE-sol1}

Considering the exponents of $t'$, we need to have $a-1 = a-2b$.  Hence $b = \frac{1}{2}$.  We can already re-write $\xi = \frac{y'}{\sqrt{t'}}$

To find the value of $a$, we will use the fact that $$\int_{-\infty}^{\infty} P(y', t') dy' = 1$$
$$\int_{-\infty}^{\infty} t'^a \cdot f \left( \frac{y'}{\sqrt{t'}} \right) dy' = 1$$

Using the substitution $u = \frac{y'}{\sqrt{t'}}$, we have $\frac{du}{dy'} = t'^\frac{-1}{2}$

$$t'^a \int_{-\infty}^{\infty} f(u) du \cdot t'^\frac{1}{2} = 1$$
$$t'^{a+\frac{1}{2}} \int_{-\infty}^{\infty} f(u) du = 1$$ 

Considering $$\int_{-\infty}^{\infty} f(u) du = 1$$, we deduce that $$a+\frac{1}{2} = 0$$ and $$a = \frac{-1}{2}$$

Re-writing @eq-FKE-sol1 using our new values for $a$ and $b$: 

$$\frac{-1}{2} f(\xi) - \frac{1}{2} \xi = C^2 \frac{d^2f}{d \xi^2}$$
$$\frac{-1}{2} \frac{d(\xi f(\xi))}{d(\xi)} = C^2 \frac{d^2f}{d \xi^2}$$
$$\frac{-1}{2} \xi f(\xi)) = C^2 \frac{df}{d \xi} + const.$$

We make the constant = 0.

$$\frac{-1}{2} \xi = C^2 \frac{1}{f(\xi)} \frac{df}{d \xi}$$
$$\frac{-1}{2} \xi = C^2 \frac{d(log \space f(\xi))}{d(f(\xi))} $$ 

Integrating both side for $\xi$, we get: 
$$\frac{-1}{2} \int \xi \space d\xi= C^2 log \space f(\xi)$$

$$\frac{-1}{2} \frac{\xi^2}{2} + c_1= C^2 log \space f(\xi)$$

$$log \space f(\xi) = \frac{-1}{4 C^2} \xi^2 + c$$

$$f(\xi) = e^{\frac{-1}{4 C^2} \xi^2 + c} = A \cdot e^{\frac{-1}{4 C^2} \xi^2}$$
Time, to revisit our initial solution @eq-FKE-sol: 

$$P(y, t; y', t') = \frac{1}{\sqrt{t}} A \cdot e^{\frac{-1}{4 C^2} \frac{y'^2}{t'}}$$

We choose $A$ such that 
$$\int_\mathbb{R} f(\xi) \space d\xi = 1$$ 

$$A \cdot \int_\mathbb{R} e^{\frac{-1}{4 C^2} \xi^2} \space d\xi = 1$$

Using substitution $x = \frac{\xi}{2C}$, we get $\frac{dx}{d\xi} = \frac{1}{2C}$, hence:

$$A \cdot 2C \cdot \int_\mathbb{R} e^{-x^2} dx = 1$$

$$A = \frac{1}{2C \sqrt{\pi}}$$
$$P(y, t; y', t') = \frac{1}{\sqrt{t}} \space \cdot \frac{1}{2C \sqrt{\pi}} \cdot e^{\frac{-1}{4 C^2} \frac{y'^2}{t'}}$$


::: {.callout-note}
# Probability Density Function for normal distribution

Recall the Probability Density Function for a random normal variable. 

$$f(x) = \frac{1}{\sigma \sqrt{2 \pi}} \cdot e^{\frac{-1}{2} \frac{(x-\mu)^2}{\sigma^2}}$$
:::

With this in mind, we could set $\sigma = C \cdot \sqrt{2t'}$ and $\sigma^2 = 2 \cdot C^2 \cdot t'$

$$P(y, t; y', t') = \frac{1}{\sigma \sqrt{2 \pi}} \space \cdot e^{\frac{-1}{2} \frac{y'^2}{\sigma^2}}$$

Hence $y'$ is a random variable such that $$y' \sim N \left( 0, (C \sqrt{2t'} \space)^2 \right)$$


# An Alternative Situation 

We could imagine that instead of the trinomial walk, we a have a **Symmetric Random Walk**.  We defined some properties of such symmetric random walk in an earlier [post](../random-behavior-assets/index.qmd)

This mean that our random walker can move in steps $\delta y$ in time-step $\delta t$.  

Similarly to what we have done above, 
$$Prob(a<y'<b \text{ at time t' } | \text{ y at time t}) = \int_a^b p(y, t; y', t') dy'$$
and 

$$P(y, t; y', t') = \alpha \cdot P(y, t; y'+\delta y, t'-\delta t) + \alpha \cdot P(y, t; y'-\delta y, t' - \delta t)$$ Because the random walk is symmetric $\alpha$ has to be $\frac{1}{2}$

$$P(y, t; y' + \delta y, t'-\delta t)  \approx P(y,t;y',t') + \delta y \frac{\partial P}{\partial y'} - \delta t \frac{\partial P}{\partial t'} + \frac{1}{2} \delta y^2 \frac{\partial^2 P}{\partial y'^2} + \dots$$

$$P(y, t; y'- \delta y, t'-\delta t)  \approx P(y,t;y',t') - \delta y \frac{\partial P}{\partial y'} - \delta t \frac{\partial P}{\partial t'} + \frac{1}{2} \delta y^2 \frac{\partial^2 P}{\partial y'^2} + \dots$$

Putting it all back together and again using $P(y, t; y', t') = P(y', t')$ for brevity: 

$$P(y', t') = P(y', t') - \delta t \frac{\partial P}{\partial t'} + \frac{1}{2} \delta y^2 \frac{\partial^2 P}{\partial y'^2}$$

$$\delta t \frac{\partial P}{\partial t'} = \frac{1}{2} \delta y^2 \frac{\partial^2 P}{\partial y'^2}$$

$$\frac{\partial P}{\partial t'} = \frac{1}{2} \frac{\delta y^2}{\delta t} \frac{\partial^2 P}{\partial y'^2}$$

When we take the limit for $\delta t$ and $\delta y'^2$ to tend -> 0, this expression would only make sense if  $\frac{\delta y^2}{\delta t} -> 1$; otherwise the probability will either be $0$ or $\infty$.  

$$\frac{\partial P}{\partial t'} = \frac{1}{2} \frac{\partial^2 P}{\partial y'^2}$$




