---
title: "04 - Time-series decomposition"
author: "François de Ryckel"
date: "2022-10-21"
categories: [Time-Series]
description: 'Introducing time-series decomposition. We first show how to compose time-series using linear trend, seasonality and then white nosie.'
editor: source
date-modified: "2023-10-22"
---

When studying time-series, we usually consider trend and seasonality.  With that in mind, we define time-series either in an additive way $$y_t = T_t + S_t + R_t$$ or in a multiplicative way $$y_t = T_t \cdot S_t \cdot R_t$$ 

with: 


* $y_t$ the value of the time-serie at time $t$
* $T_t$ the trend component at time $t$
* $S_t$ the seasonal component at time $t$
* $R_t$ the remainder component at time $t$

::: {.callout-note}

Note that, using log, we can easily transform a multiplicative time-series into an addidiative one. 

$$log(y_t) = log(T_t \cdot S_t \cdot R_t) = log(T_t) + log(S_t) + log(R_t)$$
:::

Because decomposition used averages (for seasonal) and noving average (for trend), decomposition is not robust to outlier (aka, it is very sensitive to outlier)

# Time-series composition 

Perhaps before to decompose time-series, we can try to compose some dummy time-series in both additive or multiplicative ways. 

```{r}
#| message: false
#| warning: false
#| echo: false

library(dplyr)
library(ggplot2)
```

## Compose deterministic time-series (with a trend and a stochastic component) 

```{r}
#| label: time_series_composition

df <- tibble(x = 1:252, phi = rnorm(252, mean = 0, sd = 1.5)) |> 
  mutate(y = 0.07 * x + 0.03 + phi)

ggplot(df, aes(x, y)) + 
  geom_line() + 
  ggtitle(label = 'Compose a linear trend with a stochastic component')

  
```

We could also create a seasonal time-series with a stochastic component. 

```{r}
df <- tibble(x = 1:252, phi = rnorm(252, mean = 0, sd = 1.5)) |> 
  mutate(y = 1.7 * sin((2 * pi * x / 50) + 0.3 * pi ) + phi)
         
ggplot(df, aes(x, y)) + 
  geom_line() + 
  ggtitle(label = 'Compose a seasonal trend with a stochastic component')
```

## Decomposing a time-series. 

### In R using standard library

Let's go back to the milk example which looks like this.   

```{r}
#| message: false
#| label: milk1

library(readr)

milk <- read_csv('../../../raw_data/milk.csv')

head(milk)
  
ggplot(milk, aes(x = month, y = milk_prod_per_cow_kg)) + 
  geom_line()
```

If we transform our tibble into a time-series df, then we can use the *decompose()* function.  

```{r}
ts_milk <- ts(milk$milk_prod_per_cow_kg, start = c(1962, 1, 1), frequency = 12)
milk_dec <- decompose(ts_milk, type = 'additive', filter = NULL)

plot(milk_dec)
```

And now, we can use our decomposed time-series to detrend or remove seasonality 

```{r}
milk_adj <- ts_milk - milk_dec$seasonal
plot(milk_adj)
```

The *timtk* package can achieve the same output in a more direct way as timetk fit the tidyverse framework. 

```{r}
#| fig-height: 15

library(timetk)

milk |> 
  plot_stl_diagnostics(month, milk_prod_per_cow_kg, .frequency = 12)

```

It should be noted that the *timetk* package use a Seasonal-Trend-Loess decomposition (STL).
If we want to get the values, we use the **tk_stl_diagnostics(date, value)** function. 

([From FPP 3rd ed.](https://otexts.com/fpp3/stl.html) )While loess is a method for estimating nonlinear relationships. The STL method was developed by R. B. Cleveland et al. (1990).

STL has several advantages over classical decomposition:

* STL will handle any type of seasonality, not only monthly and quarterly data.
* The seasonal component is allowed to change over time, and the rate of change can be controlled by the user.
* The smoothness of the trend-cycle can also be controlled by the user.
* It can be robust to outliers (i.e., the user can specify a robust decomposition), so that occasional unusual observations will not affect the estimates of the trend-cycle and seasonal components. They will, however, affect the remainder component. 


```{r}
head(milk |> tk_stl_diagnostics(month, milk_prod_per_cow_kg))
```

## Moving averages 

There are so many packet in R to calculate moving averages.  We'll keep using the timetk package 

```{r}
yo <- milk |> 
  mutate(ma_3m = slidify_vec(.x = milk_prod_per_cow_kg, .f = mean, .period = 3, .align = 'right'))
```

